---
title: "M3_W3_06_Managing_Dataframes_With_DPLYR_-_Basic_Tools"
author: "Alexander Cormack"
date: "3 September 2022"
output: html_document
---

# Managing Dataframes With DPLYR - Basic Tools

For this lecture we will be using a dataset of daily pollution measurements in Chicago from 1987 to 2005.

```{r}
library(dplyr)

chicago <- readRDS("./data/chicago.rds")
str(chicago)
```


### Select

The select function is used to view specific columns of a dataframe.

We can look at the names of a dataframe with the names function. This is useful because one of the nice things about the dplyr package is that you can access the column variables with the select function using the column names and normal indexing notation - you don't need to use the column indices:

```{r}
names(chicago)
```
```{r}
head(select(chicago, city:date))
```



We can also use the minus sign to view all of the columns except the ones we indicate:

```{r}
head(select(chicago, -(city:date)))
```


To do this outside of the dplyr package you would need to find the index of the city column, the index of the date column and then use standard slicing methods:

```{r}
i <- match("city", names(chicago))
j <- match("date", names(chicago))
head(chicago[, -(i:j)])
```



### Filter

The filter function is used to subset rows of a dataframe based on particular conditions. So for example we might want to see all the rows of the Chicago dataset where pm25tmean2 is greater than 30:

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30)
head(chic.f)
```



We could also create a more complicated logical sequence my creating multiple conditions:

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80)
head(chic.f, 10)
```


### Arrange

Arrange is used to reorder the rows of a dataframe based on the values of a column. For example, we could arrange the dataframe by date so that the oldest date appears first and the most recent date appears last:

```{r}
chicago <- arrange(chicago, date)
head(chicago, 3)
tail(chicago, 3)
```


Of course we could arrange the dates in descending order as follows:

```{r}
chicago <- arrange(chicago, desc(date))
head(chicago, 3)
tail(chicago, 3)
```



### Rename

This is a simple function that allows you to rename a variable. It is however very useful because if you don't have this function renaming variables is a surprisingly difficult thing to do in R.

Some of the variables in the chicago dataset are a bit of a mouthful and not very intuitive (e.g. pm25tmean2 or dptp).So we can use the function by passing the dataframe name, and then the new name and the old name:

```{r}
chicago <- rename(chicago, pm25 = pm25tmean2, dewpoint = dptp)
head(chicago)
```



### Mutate

The mutate function is used to transform existing variables or create new variables. Here we create a new variable called pm25detrend which is the pm25 variable with the mean subtracted from it (aim is to show deviations from the mean).

```{r}
chicago <- mutate(chicago, pm25detrend = pm25 - mean(pm25, na.rm = TRUE))
head(select(chicago, pm25, pm25detrend))
```


### Group_by

This function allows you to split a dataframe by certain categorical variables. In this example we create a temperature categorical variable that will indicate whether a given day is hot or cold. So first we create a factor variable with a logical operator (a hot day is greater than 80 degrees F). Then we use the group_by function to create a new data structure that has a slightly different format: 

```{r}
chicago <- mutate(chicago, tempcat = factor(1 * (tmpd > 80), labels = c("cold", "hot")))
hotcold <- group_by(chicago, tempcat)
hotcold
```



We could now use the summarise function on our new data structure to get some more insights about the data distribution on cold and hot days. Let's say we want to know what the mean value of pm25, the maximum value of ozone and the median value of nitrogen on both hot and cold days:

```{r}
summarise(hotcold, pm25 = mean(pm25, na.rm = TRUE), o3 = max(o3tmean2), no2 = median(no2tmean2))
```







