---
title: "M3_W3_01_Subsetting_and_Sorting"
author: "Alexander Cormack"
date: "31 August 2022"
output: html_document
---

# Subsetting - quick review

We now have learnt how to load data ... now we want to manipulate that data and set it up as a tidy dataset.

To practice this, let's create a dataframe and scramble the variables

```{r}
set.seed(13435)
X <- data.frame("var1" = sample(1:5), "var2" = sample(6:10), "var3" = sample(11:15))
X <- X[sample(1:5), ]; X$var2[c(1, 3)] = NA
X
```



First thing we can do is subset a specific column by the column index:

```{r}
X[, 1]
```


I can do the same by using the column name:

```{r}
X[, "var1"]
```


I can also subset by both rows and columns at the same time by mixing row indices and column names:

```{r}
X[1:2, "var2"]
```



We can also subset using logicals 'ands' and 'ors'. Let's say we want to subset where var1 is less than or equal to 3
and var3 is greater than 11:

```{r}
X[(X$var1 <= 3 & X$var3 > 11), ]
```


Here we want all rows where var1 is less than or equal to 3 or var3 is greater than 12:

```{r}
X[(X$var1 <= 3 | X$var3 > 12), ]
```


I can deal with NA values by using the 'which' command to filter them out:

```{r}
X[which(X$var2 > 8), ]
```


# Sorting







# CLEAN UP #################################################

# Clear environment
rm(list = ls()) 

# Clear packages
p_unload(all)  # Remove all add-ons

# Clear plots
dev.off()  # But only if there IS a plot

# Clear console
cat("\014")  # ctrl+L

# Clear mind :)
