---
title: "M3_W3_01_Subsetting_and_Sorting"
author: "Alexander Cormack"
date: "31 August 2022"
output: html_document
---

# Subsetting - quick review

We now have learnt how to load data ... now we want to manipulate that data and set it up as a tidy dataset.

To practice this, let's create a dataframe and scramble the variables

```{r}
set.seed(13435)
X <- data.frame("var1" = sample(1:5), "var2" = sample(6:10), "var3" = sample(11:15))
X <- X[sample(1:5), ]; X$var2[c(1, 3)] = NA
X
```



First thing we can do is subset a specific column by the column index:

```{r}
X[, 1]
```


I can do the same by using the column name:

```{r}
X[, "var1"]
```


I can also subset by both rows and columns at the same time by mixing row indices and column names:

```{r}
X[1:2, "var2"]
```



We can also subset using logicals 'ands' and 'ors'. Let's say we want to subset where var1 is less than or equal to 3
and var3 is greater than 11:

```{r}
X[(X$var1 <= 3 & X$var3 > 11), ]
```


Here we want all rows where var1 is less than or equal to 3 or var3 is greater than 12:

```{r}
X[(X$var1 <= 3 | X$var3 > 12), ]
```


I can deal with NA values by using the 'which' command to filter them out:

```{r}
X[which(X$var2 > 8), ]
```


# Sorting

The sort command will sort variables - the default is in increasing order. You can also put NA values at the end

```{r}
sort(X$var1)

sort(X$var1, decreasing = TRUE)

sort(X$var2, na.last = TRUE)
```


# Ordering

We can also order a dataframe by a particular variable. Let's say we want to sort all of the values in var1
in ascending order and order the values in var2 and var3 accordingly, i.e. sort the rows by the values in var1

```{r}
X[order(X$var1), ]
```



You can also order by multiple variables. So if in particular rows there are identical values in the two different variables it will sort them accordingly (N.B. there are no identical variables in the example below)

```{r}
X[order(X$var1, X$var3), ]
```


# Ordering with plyr

The same can be done with the plyr package and the arrange command. The arrange command takes as arguments a dataframe and a variable. The default is ascending order, but of course there is a command for descending order

```{r}
library(plyr)
arrange(X, var1)

arrange(X, desc(var1))
```



# Adding rows and columns

This is quite straightforward. In the example below a vector (in this case a series of random normals) is assigned to a new column (var4)

```{r}
X$var4 <- rnorm(5)
X
```


Another way you can do this is by using the cbind command. In the example below the vector rnorm(5) is bound as a column to the right side of X. Inverting the order of the dataframe and the vector will bind the column to the left side of the dataframe

```{r}
Y <- cbind(X, rnorm(5))
Y
```



Rows can be added by using the rbind command. Analogously, if place the vector to the right of the dataframe rbind will bind the vector to the bottom of the dataframe, and if you put it to the left it will bind it to the top of the dataframe

```{r}
Z <- rbind(Y, rnorm(5))
Z
```

