---
title: "M3_W2_02_Reading_from_HDF5"
author: "Alexander Cormack"
date: "22 August 2022"
output: html_document
---

### HDF5

* Used for storing large datasets
* Supports storing a range of data types
* Heirarchical data format
* Groups containing zero or more datasets and metadata
    + have a group header with group name and list of attributes
    + have a group symbol table with a list of objects in the group
* Datasets are multidimensional arrays of data elements with metadata that
    + have a header with name, datatype, dataspace, and storage layout
    + have a data array with the data

http://www.hdfgroup.org/



### R HDF5 package

This is installed from Bioconductor http://bioconductor.org/ (primarily used for genomics but also has good 'big data' packages)

It can be used to interface with hdf5 datasets

First we need to check whether we need to install the BiocManager package (this is used to install packages from Bioconductor)

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install()
```


Then we can install the rhdf5 package with BiocManager

```{r}
BiocManager::install(c("rhdf5"))
```


Now let's load the package and create a file

```{r}
library(rhdf5)
created = h5createFile("example.h5")
created
```



### Create groups

Now we can create groups - something like a folder strucuture - and list them

```{r}
created = h5createGroup("example.h5", "foo")
created = h5createGroup("example.h5", "baa")
created = h5createGroup("example.h5", "foo/foobaa")
h5ls("example.h5")
```



### Write to groups

We can use the h5write command to write a variety of different objects to specific groups in a file


```{r}
A = matrix(1:10, nr = 5, nc = 2)
h5write(A, "example.h5", "foo/A")
B = array(seq(0.1, 2.0, by = 0.1), dim = c(5, 2, 2))
attr(B, "scale") <- "liter"
h5write(B, "example.h5", "foo/foobaa/B")
h5ls("example.h5")
```


### Write a dataset

```{r}
df = data.frame(1L:5L, seq(0, 1, length.out = 5), 
                c("ab", "cde", "fghi", "a", "s"), stringsAsFactors = FALSE)
h5write(df, "example.h5", "df")
h5ls("example.h5")
```



### Reading data

```{r}
readA = h5read("example.h5", "foo/A")
readB = h5read("example.h5", "foo/foobaa/B")
readdf = h5read("example.h5", "df")
readA
readB
readdf
```



### Writing and reading in chunks

```{r}
h5write(c(12, 13, 14), "example.h5", "foo/A", index = list(1:3, 1))
h5read("example.h5", "foo/A")
```



```{r}
h5read("example.h5", "foo/A", index = list(1:3, 1))
```














