---
title: "M3_W4_01 Editing_text_variables"
author: "Alexander Cormack"
date: "2022-10-12"
output: html_document
---

# Getting and cleaning data

This lecture uses Baltimore City traffic camera data ... but the dataset is not available!!!

### Fixing character vectors - tolower(), toupper()

You can use these commands to convert all characters to lower or upper case
Take for example column names that sometimes appear with mixed case. This can be annoying if you have to constantly input the column names.

```{r}
col_names <- c("address", "direction", "street", "crossStreet", "intersection", "Location.1")
tolower(col_names)
```

### Fixing character vectors - strsplit()

This command can be useful for splitting strings by some defined character.

The important parameters are: x, split

```{r}
split_names <- strsplit(col_names, "\\.")
split_names
```

### Quick aside - lists

```{r}
my_list <- list(letters = c("A", "b", "c"), numbers = 1:3, matrix(1:25, ncol = 5))
head(my_list)
```

We can select the first element of the list:

```{r}
my_list[1]
```

We can subset by naming variables:

```{r}
my_list$letters
```

We can also subset the vector by taking the first element of the list:

```{r}
my_list[[1]]
```

### Fxing character vectors - sapply()

sapply() applies a function to each element in a vector or list.

The important parameters are: X, FUN

Say for example we removed all of the full stops from the column names and now we only want the first element of the names.

We can write a function to return only the first element and use sapply() to apply it to all of the column names:

```{r}
first_element <- function(x){x[1]}
sapply(split_names, first_element)
```

Now let's play around with the Peer review data from a previous lecture

```{r}
reviews <- read.csv("./data/reviews.csv")
solutions <- read.csv("./data/solutions.csv")

head(reviews, 2)
```

```{r}
head(solutions, 2)
```


### Fixing character vectors - sub()

The sub() command substitutes characters in a string.

Say for example we would like to remove the underscores in all of the column names:

```{r}
names(reviews)
```

```{r}
sub("_", "", names(reviews), )
```

### Fixing character vectors - gsub()

gsub() works like sub, only that it will substitute all of the instances of that character in a string.

Compare the difference in the two blocks of code below:

```{r}
test_name <- "this_is_a_test"
sub("_", "", test_name)
```

```{r}
gsub("_", "", test_name)
```

### Finding values - grep(), grepl()

grep() will search for a within a specific variable.


grep("searchTerm", dataFrame$columnName)

[1] 4 5 36  (i.e. the searchTerm appears in the 4th, 5th and 36th element of the columnName variable)


table(grepl("searchTerm", dataFrame$columnName))

FALSE   TRUE
   77      3


If you use the value=TRUE parameter grep() will return the value rather than the position

grep("searchTerm", dataFrame$columnName, value=TRUE)

[1] "searchTerm in the string"   "This is the searchTerm"   "The searchTerm again"


If you want to be sure a value doesn't appear in a varialbe you can grep() the value and call the length of the result

length(grep("someOtherTerm", dataFrame$columnName))

[1] 0
   
### More useful string functions

nchar will tell you the number of characters in a string

```{r}
nchar("Alexander Cormack")
```

You can subset a string by calling on a series of indices

```{r}
substr("Alexander Cormack", 1, 9)
```

You can paste two strings together with the paste function

```{r}
paste("Alexander", "Cormack")
```

You can set the character that you want to separate the pasted strings with the sep parameter

```{r}
paste("Alexander", "Cormack", sep = "$")
```



If you want to paste strings together with no separator then you can use the paste0 function

```{r}
paste0("Alexander", "Cormack")
```

You can trim excess whitespace with the str_trim function from the stringr package

```{r}
library(stringr)
str_trim("     Alexander        ")
```

### Some important points about text in datasets

- Names of variables should be
    all lower case when possible
    descriptive (Diagnosis vs Dx)
    not duplicated
    not have underscores or dots or whitespaces

- Variables with character values
    should usually be made into factor variables (depends on application)
    should be descriptive (use TRUE/FALSE instead of 0//1 and Male/Female versus 0/1 or M/F)