---
title: "M3_W2_04_Application_programming_interfaces"
author: "Alexander Cormack"
date: "22 August 2022"
output: html_document
---

N.B. Couldn't manage to get things to work for this unit. Have bought Russell & Klassen 'Mining the Social Web 3E' to delve into it further


https://developer.twitter.com/en/docs/tutorials/getting-started-with-r-and-v2-of-the-twitter-api


### Getting started with the Twitter API

We'll be using the httr package to get website information.

In order to get information from a website via an API first you will need to have a developer account. Here I will be accessing Twitter, so I have set up a developer account and I now have a bearer token.



### Accessing Twitter from R

We are using the httr package and the command oauth_app - this starts the authorisation process. We then use the GET command and the url corresponding to the Twitter API (version 1.1). Then we pass in the statuses on the home timeline which is delivered as a JSON file. Lastly we pass the authentication with the oauth sign in.

```{r}
# myapp = oauth_app("twitter", key = "yourAPIKeyHere", secret = "yourAPIKeySecretHere")
# sig = sign_oauth1.0(myapp, token = "yourAccessTokenHere", token_secret = "yourAccessTokenSecretHere")
# homeTL = GET("https://api.twitter.com/1.1/statuses/home_timeline.json", sig)
```


### Converting the JSON object



```{r}
library(rjson)

json1 = content(homeTL)
json2 = jsonlite::fromJSON(toJSON(json1))
json2
```





